Split-Apply-Combine with Dynamic Grouping

### 2024-03-27 paper/article.Rnw
changed verbatim to Code environment

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5%%%%%%%
\section[R package accumulate]{\proglang{R Package} \pkg{accumulate} }\label{sect:accumulate}
Grouped aggregation with a fall-though scenario based on a collapsing scheme
requires a fair amount of specification by the user. Besides the data to be
aggregated, one needs to specify the method(s) of aggregation, the collapsing
scheme, and the condition to decide whether a subset is fit for aggregation or
a next collapse is necessary. There are two main functions in \pkg{accumulate}
that offer slightly different interfaces.
\begin{verbatim}
  accumulate(data, collapse, test, fun, ...)
  cumulate(data, collapse, test, ...)
\end{verbatim}
Here \texttt{data} is a data frame holding data to be aggregated;
\texttt{collapse} represents the collapse sequence (as a \class{formula} or a
\class{data frame}), and \texttt{test} is a function that accepts a subset of
\texttt{data} and returns a boolean that indicates whether a subset is suited
for aggregation or not. In \code{accumulate()}, the parameter \code{fun}
represents an aggregation function that is applied to every column of
\code{data}, and the ellipsis (\code{...}) is for arguments that are passed as
extra argument to \code{fun}. The interface of \code{accumulate()} is somewhat
similar to that of the \code{aggregate()} function in \proglang{R}.  In
\code{cumulate()}, the ellipsis is a sequence of comma-separated
\code{name = expression} pairs in the style of \code{summarise()} from the
\code{dplyr} package.

The output of both functions are of the same form. The columns of the output
data frame and can schematically be represented as follows. 
\begin{verbatim}
  [Grouping Variables, Collapse level, Output aggregates]
\end{verbatim}
The first columns represent the variables that define the output grouping, the
next column is an integer that indicates the level of collapsing used to
compute the aggregate (0 indicating no collapse), and the last set of columns
store the aggregates. Output aggregates may be of a simple data type
(\code{numeric}, \code{character}, \code{logical},$\ldots$) or of a composed
type such as the output of a linear model. The latter case is demonstrated in
Section~\ref{sect:complex}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The Formula Interface}\label{sect:formula}
We will use the example of Section~\ref{sect:dynamicgrouping} to illustrate
the \class{formula} interface.
<<>>=
library("accumulate")
input <- data.frame(
     A  = c( 1,  1,  1,  2,  2,  2,  3,  3,  3)
   , B  = c(11, 11, 11, 12, 12, 13, 21, 22, 12)
   , B1 = c( 1,  1,  1,  1,  1,  1,  2,  2,  1)
   , Y  = 1:9
)

@@

Summarizing, the \class{formula} interface is always of the following form.
\begin{verbatim}
  Target grouping ~ Alternative1 + Alternative2 + ... + AlternativeN
\end{verbatim}

@@

It is interesting to see how the elements $U$, $f$, and $\phi$ are implemented
in practice. Consider the signature of the \proglang{R}'s \code{aggregate()}
function (we skip arguments that are not important for the discussion).
\begin{verbatim}
  aggregate(x, by, FUN)
\end{verbatim}
Here, \code{x} is a data frame where each row represents an element of $U$. The
parameter \code{by} is a list of vectors of group labels, where each vector has
a length that equals the number of rows in \code{x}. So the function $f:U\onto
A$ is implemented by asking the user to make sure that the position of each
label in \code{by} corresponds to the correct row number in the data frame
\code{x}. The argument \code{FUN} (of class \class{function}) represents the
function $\phi$ that aggregates each subset of \code{x}. When the records in
\code{x} contain more than one variable, the aggregator is applied to each one
of them.  Here is an example of how a user might call this function from the
\proglang{R} prompt.
%
<<>>=
aggregate(iris[1:2], by = iris["Species"], FUN = mean)
@
%
Note that the correspondence in position of the \texttt{Species} label and the
record position is implemented by taking them from the same data frame. The
output also reveals in the first column the set $A$: each row corresponds to a
unique value in \texttt{Species} column.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5%%%%%%%
\section[R Package accumulate]{\proglang{R Package} \pkg{accumulate} }\label{sect:accumulate}
Grouped aggregation with a fall-though scenario based on a collapsing scheme
requires a fair amount of specification by the user. Besides the data to be
aggregated, one needs to specify the method(s) of aggregation, the collapsing
scheme, and the condition to decide whether a subset is fit for aggregation or
a next collapse is necessary. There are two main functions in \pkg{accumulate}
that offer slightly different interfaces.
\begin{Code}
  accumulate(data, collapse, test, fun, ...)
  cumulate(data, collapse, test, ...)
\end{Code}
Here \texttt{data} is a data frame holding data to be aggregated;
\texttt{collapse} represents the collapse sequence (as a \class{formula} or a
\class{data frame}), and \texttt{test} is a function that accepts a subset of
\texttt{data} and returns a boolean that indicates whether a subset is suited
for aggregation or not. In \code{accumulate()}, the parameter \code{fun}
represents an aggregation function that is applied to every column of
\code{data}, and the ellipsis (\code{...}) is for arguments that are passed as
extra argument to \code{fun}. The interface of \code{accumulate()} is somewhat
similar to that of the \code{aggregate()} function in \proglang{R}.  In
\code{cumulate()}, the ellipsis is a sequence of comma-separated
\code{name = expression} pairs in the style of \code{summarise()} from the
\code{dplyr} package.

The output of both functions are of the same form. The columns of the output
data frame and can schematically be represented as follows. 
\begin{Code}
  [Grouping Variables, Collapse level, Output aggregates]
\end{Code}
The first columns represent the variables that define the output grouping, the
next column is an integer that indicates the level of collapsing used to
compute the aggregate (0 indicating no collapse), and the last set of columns
store the aggregates. Output aggregates may be of a simple data type
(\code{numeric}, \code{character}, \code{logical},$\ldots$) or of a composed
type such as the output of a linear model. The latter case is demonstrated in
Section~\ref{sect:complex}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The Formula Interface}\label{sect:formula}
We will use the example of Section~\ref{sect:dynamicgrouping} to illustrate
the \class{formula} interface.
<<>>=
library("accumulate")
input <- data.frame(
  A  = c( 1,  1,  1,  2,  2,  2,  3,  3,  3),
  B  = c(11, 11, 11, 12, 12, 13, 21, 22, 12),
  B1 = c( 1,  1,  1,  1,  1,  1,  2,  2,  1),
  Y  = 1:9
)

@@

Summarizing, the \class{formula} interface is always of the following form.
\begin{Code}
  Target grouping ~ Alternative1 + Alternative2 + ... + AlternativeN
\end{Code}

@@

It is interesting to see how the elements $U$, $f$, and $\phi$ are implemented
in practice. Consider the signature of the \proglang{R}'s \code{aggregate()}
function (we skip arguments that are not important for the discussion).
\begin{Code}
  aggregate(x, by, FUN)
\end{Code}
Here, \code{x} is a data frame where each row represents an element of $U$. The
parameter \code{by} is a list of vectors of group labels, where each vector has
a length that equals the number of rows in \code{x}. So the function $f:U\onto
A$ is implemented by asking the user to make sure that the position of each
label in \code{by} corresponds to the correct row number in the data frame
\code{x}. The argument \code{FUN} (of class \class{function}) represents the
function $\phi$ that aggregates each subset of \code{x}. When the records in
\code{x} contain more than one variable, the aggregator is applied to each one
of them.  Here is an example of how a user might call this function from the
\proglang{R} prompt.
%
<<>>=
aggregate(iris[1:2], by = iris["Species"], FUN = mean)
@
%
Note that the correspondence in position of the \texttt{Species} label and the
record position is implemented by taking them from the same data frame. The
output also reveals in the first column the set $A$: each row corresponds to a
unique value in \texttt{Species} column.
